---
title: "TDDE15-Lab 1"
author: "Fredrik Ramberg"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

# 1

-   Healthy –\> healthy: 0.9

-   Infected -\> infected: 0.8

-   p(Test = pos \| infected ) = 0.6

-   p(Test = negative \| healthy) = 0.7

Question: p(healthy in three days\| negative test third day)

```{r}
library(bnlearn)
library(gRain)

dag = model2network("[H1][H2|H1][H3|H2][T1|H1][T2|H2][T3|H3]")
graphviz.plot(dag)

# 1 if healthy
H1 <- c(0.5, 0.5)
dim(H1) <- c(2)
dimnames(H1) <- list(c("0", "1"))


#####  New
# Old
H2 <- matrix(c(0.8, 0.2,
               0.1, 0.9), nrow = 2, ncol = 2)
dim(H2) <- c(2,2)
dimnames(H2) <- list("H2" = c("0", "1"), "H1" = c("0", "1"))

H3 <- matrix(c(0.8, 0.2,
               0.1, 0.9), nrow = 2, ncol = 2)
dim(H3) <- c(2,2)
dimnames(H3) <- list("H3" = c("0", "1"), "H2" = c("0", "1"))

# 1 if testing positive
T1 <- matrix(c(0.4, 0.6,
               0.7, 0.3), nrow = 2, ncol = 2)
dim(T1) <- c(2,2)
dimnames(T1) <- list("T1" = c("0", "1"), "H1" = c("0", "1"))

T2 <- matrix(c(0.4, 0.6,
               0.7, 0.3), nrow = 2, ncol = 2)
dim(T2) <- c(2,2)
dimnames(T2) <- list("T2" = c("0", "1"), "H2" = c("0", "1"))

T3 <- matrix(c(0.4, 0.6,
               0.7, 0.3), nrow = 2, ncol = 2)
dim(T3) <- c(2,2)
dimnames(T3) <- list("T3" = c("0", "1"), "H3" = c("0", "1"))

bn <- custom.fit(dag, list(H1=H1, H2=H2, H3=H3, T1=T1, T2=T2, T3=T3))

bn_comp <- compile(as.grain(bn))
```

```{r}
nodes <- c("T2")#, "T3")
states <- c("0")#, "0")

querygrain(bn_comp, c("H3"))
querygrain(setEvidence(bn_comp,nodes=c("T2"),states=c("0")),c("H3"))
querygrain(setEvidence(bn_comp,nodes=c("T2", "T3"),states=c("0", "0")),c("H3"))

```

-   The probability that she is healthy after 3 days is 58.5%

-   The probability that she is healthy after three days given a negative test on the second day is 68%

-   The probability that she is healthy after three days given negative test on the second and third day is 79%.

# 2

-   p(rainy today \| rainy last two days) = 0.75

    -   ((p(sunny today \| rainy last two days) = 0.25))

-   p( rainy today\| 1 rainy and 1 sunny) = 0.5

    -   ((p( sunny today\| 1 rainy and 1 sunny) = 0.5))

-   Weather forecast: reports rainy weather when the forecast is actually sunny and vice versa with probability 0.1

+-------------+-------------+-------------+-------------+-------------+
| NEW -\>\    | RR          | RS          | SR          | SS          |
| Old         |             |             |             |             |
|             |             |             |             |             |
| \|          |             |             |             |             |
+=============+=============+=============+=============+=============+
| RR          | 0.75        | 0.25        | 0           | 0           |
+-------------+-------------+-------------+-------------+-------------+
| RS          | 0           | 0           | 0.5         | 0.5         |
+-------------+-------------+-------------+-------------+-------------+
| SR          | 0.5         | 0.5         | 0           | 0           |
+-------------+-------------+-------------+-------------+-------------+
| SS          | 0           | 0           | 0.25        | 0.75        |
+-------------+-------------+-------------+-------------+-------------+
|             |             |             |             |             |
+-------------+-------------+-------------+-------------+-------------+

```{r}
library(HMM)
transProbs <- c()

#First letter is one day ago, second letter is today
states <- c("RR", "RS", "SR", "SS")
symbols <- c("R", "S")

######## Symbols
# States
emissionProbs <- matrix(c(0.9, 0.1,
                          0.1,0.9,
                          0.9, 0.1,
                          0.1,0.9), ncol=2, byrow = TRUE)
colnames(emissionProbs) <- symbols
rownames(emissionProbs) <- states


###   Old
# New
transProbs <- matrix(c(0.75, 0.25, 0, 0,
                        0, 0, 0.5, 0.5,
                        0.5, 0.5, 0, 0,
                        0, 0, 0.25, 0.75), 
                        nrow = 4, ncol = 4, byrow = TRUE)
colnames(transProbs) <- states
rownames(transProbs) <- states

emissionProbs
transProbs

hmm <- initHMM(states,symbols, transProbs = transProbs, emissionProbs = emissionProbs)

set.seed(1234)
sim <- simHMM(hmm, 10)
sim

```

# 3

**From lab 3:**

```{r}
library(ggplot2)
arrows <- c("^", ">", "v", "<")
action_deltas <- list(c(1,0), # up
                      c(0,1), # right
                      c(-1,0), # down
                      c(0,-1)) # left

vis_environment <- function(iterations=0, epsilon = 0.5, alpha = 0.1, gamma = 0.95, beta = 0){
  
  # Visualize an environment with rewards. 
  # Q-values for all actions are displayed on the edges of each tile.
  # The (greedy) policy for each state is also displayed.
  # 
  # Args:
  #   iterations, epsilon, alpha, gamma, beta (optional): for the figure title.
  #   reward_map (global variable): a HxW array containing the reward given at each state.
  #   q_table (global variable): a HxWx4 array containing Q-values for each state-action pair.
  #   H, W (global variables): environment dimensions.
  
  df <- expand.grid(x=1:H,y=1:W)
  foo <- mapply(function(x,y) ifelse(reward_map[x,y] == 0,q_table[x,y,1],NA),df$x,df$y)
  df$val1 <- as.vector(round(foo, 2))
  foo <- mapply(function(x,y) ifelse(reward_map[x,y] == 0,q_table[x,y,2],NA),df$x,df$y)
  df$val2 <- as.vector(round(foo, 2))
  foo <- mapply(function(x,y) ifelse(reward_map[x,y] == 0,q_table[x,y,3],NA),df$x,df$y)
  df$val3 <- as.vector(round(foo, 2))
  foo <- mapply(function(x,y) ifelse(reward_map[x,y] == 0,q_table[x,y,4],NA),df$x,df$y)
  df$val4 <- as.vector(round(foo, 2))
  foo <- mapply(function(x,y) 
    ifelse(reward_map[x,y] == 0,arrows[GreedyPolicy(x,y)],reward_map[x,y]),df$x,df$y)
  df$val5 <- as.vector(foo)
  foo <- mapply(function(x,y) ifelse(reward_map[x,y] == 0,max(q_table[x,y,]),
                                     ifelse(reward_map[x,y]<0,NA,reward_map[x,y])),df$x,df$y)
  df$val6 <- as.vector(foo)
  
  print(ggplot(df,aes(x = y,y = x)) +
          scale_fill_gradient(low = "white", high = "green", na.value = "red", name = "") +
          geom_tile(aes(fill=val6)) +
          geom_text(aes(label = val1),size = 4,nudge_y = .35,na.rm = TRUE) +
          geom_text(aes(label = val2),size = 4,nudge_x = .35,na.rm = TRUE) +
          geom_text(aes(label = val3),size = 4,nudge_y = -.35,na.rm = TRUE) +
          geom_text(aes(label = val4),size = 4,nudge_x = -.35,na.rm = TRUE) +
          geom_text(aes(label = val5),size = 10) +
          geom_tile(fill = 'transparent', colour = 'black') + 
          ggtitle(paste("Q-table after ",iterations," iterations\n",
                        "(epsilon = ",epsilon,", alpha = ",alpha,"gamma = ",gamma,", beta = ",beta,")")) +
          theme(plot.title = element_text(hjust = 0.5)) +
          scale_x_continuous(breaks = c(1:W),labels = c(1:W)) +
          scale_y_continuous(breaks = c(1:H),labels = c(1:H)))
  
}
```

```{r}
GreedyPolicy <- function(x, y){
  
  # Get a greedy action for state (x,y) from q_table.
  #
  # Args:
  #   x, y: state coordinates.
  #   q_table (global variable): a HxWx4 array containing Q-values for each state-action pair.
  # 
  # Returns:
  #   An action, i.e. integer in {1,2,3,4}.
  
  # Your code here.
  max <- c(which(q_table[x,y,] == max(q_table[x,y,])))
  if(length(max) > 1) {
    max <- sample(max, size = 1)
  }
  
  return (max)
  
  # max <- which.max(q_table[x,y,])#which(q_table[x,y,] == max(q_table[x,y,]))
  # 
  # return (max)#sample(max, size = 1))
  
}

EpsilonGreedyPolicy <- function(x, y, epsilon){
  
  # Get an epsilon-greedy action for state (x,y) from q_table.
  #
  # Args:
  #   x, y: state coordinates.
  #   epsilon: probability of acting randomly.
  # 
  # Returns:
  #   An action, i.e. integer in {1,2,3,4}.
  
  # Your code here.
  
  max <- GreedyPolicy(x,y)
  
  epsilon_prob <- runif(1)
  if (epsilon_prob <= epsilon){
    max <-sample(1:4,1)
  }
  
  return(max)
  # If i want to put in a validity constraint
  # repeat{
  #   new_state <- c(x,y) + unlist(action_deltas[max])
  #   xn <- new_state[1]
  #   yn <- new_state[2]
  #   if(!(yn < 1 || xn < 1 || yn > W || xn > H)){
  #     return(max)
  #   }
  # }

}

transition_model <- function(x, y, action, beta){
  
  # Computes the new state after given action is taken. The agent will follow the action 
  # with probability (1-beta) and slip to the right or left with probability beta/2 each.
  # 
  # Args:
  #   x, y: state coordinates.
  #   action: which action the agent takes (in {1,2,3,4}).
  #   beta: probability of the agent slipping to the side when trying to move.
  #   H, W (global variables): environment dimensions.
  # 
  # Returns:
  #   The new state after the action has been taken.
  
  delta <- sample(-1:1, size = 1, prob = c(0.5*beta,1-beta,0.5*beta))
  final_action <- ((action + delta + 3) %% 4) + 1
  foo <- c(x,y) + unlist(action_deltas[final_action])
  foo <- pmax(c(1,1),pmin(foo,c(H,W)))
  
  return (foo)
}


q_learning <- function(start_state, epsilon = 0.5, alpha = 0.1, gamma = 0.95, 
                       beta = 0, test = FALSE){
  
  # Perform one episode of Q-learning. The agent should move around in the 
  # environment using the given transition model and update the Q-table.
  # The episode ends when the agent reaches a terminal state.
  # 
  # Args:
  #   start_state: array with two entries, describing the starting position of the agent.
  #   epsilon (optional): probability of acting randomly.
  #   alpha (optional): learning rate.
  #   gamma (optional): discount factor.
  #   beta (optional): slipping factor.
  #   reward_map (global variable): a HxW array containing the reward given at each state.
  #   q_table (global variable): a HxWx4 array containing Q-values for each state-action pair.
  # 
  # Returns:
  #   reward: reward received in the episode.
  #   correction: sum of the temporal difference correction terms over the episode.
  #   q_table (global variable): Recall that R passes arguments by value. So, q_table being
  #   a global variable can be modified with the superassigment operator <<-.
  
  # Your code here.
  
  # h <- sample(1:nrow(reward_map), size=1)
  # w <- sample(1:ncol(reward_map), size=1)
  # state <- c(h,w)
  
  # if (test)
  #   epsilon = 0
  state <- start_state
  episode_correction <- 0
  tot_reward <- 0
  repeat{
    # Follow policy, execute action, get reward.
    action <- EpsilonGreedyPolicy(x = state[1], y = state[2], epsilon = epsilon)
    next_state <- transition_model(x = state[1], y = state[2], action, beta)
    reward <- reward_map[next_state[1], next_state[2]]
    
    if (!test){
      temp_diff <- alpha*(reward + gamma * max(q_table[next_state[1], next_state[2],]) - q_table[state[1], state[2], action])
    q_table[state[1], state[2], action] <<- q_table[state[1], state[2], action] + temp_diff
    
    episode_correction <- episode_correction + temp_diff
    }
    tot_reward <- tot_reward + reward
    # Q-table update.
  
    if(reward!=0)
      # End episode.
      return (c(tot_reward,episode_correction))
    state <- next_state
  }
}
```

```{r fig.height = 7}

# Environment B (the effect of epsilon and gamma)

H <- 7
W <- 8


reward_map <- matrix(0, nrow = H, ncol = W)
reward_map[1,] <- -1
reward_map[7,] <- -1
reward_map[4,5] <- 5
reward_map[4,8] <- 10

MovingAverage <- function(x, n){
  
  cx <- c(0,cumsum(x))
  rsum <- (cx[(n+1):length(cx)] - cx[1:(length(cx) - n)]) / n
  
  return (rsum)
}

epsilons <- c(0.1, 0.25, 0.5)
gammas <- c(0.5, 0.75, 0.95)

avg_reward <- matrix(NA, ncol = length(epsilons), nrow = length(gammas), dimnames = list("epsilon" = epsilons, "gamma" = gammas))

for (e in 1: length(epsilons)){
  for (g in 1: length(gammas)){
    q_table <- array(0,dim = c(H,W,4))
    for(i in 1:30000){
      foo <- q_learning(epsilon = epsilons[e], gamma = gammas[g], start_state = c(4,1))
      # reward <- c(reward,foo[1])
      # correction <- c(correction,foo[2])
    }
    reward <- 0
    for ( i in 1:1000){
       foo <- q_learning(epsilon = epsilons[e],test = TRUE, gamma = gammas[g], start_state = c(4,1))
       reward <- reward + foo[1]
    }
    avg_reward[e, g] <- reward/1000
    
  }
}

avg_reward

#q_learning(test = TRUE, gamma = gammas[1], start_state = c(4,1))
```

```{r}

```
