## Probabalistic models 

the algorithm returns the markov blanket of Z. The markov blanket is the smallest set of nodes that makes Z conditionally interdependent of all other nodes in the graph. This means that given the markov blanket Z has no dependencies outside of the set. 

set out to empty set

loop while Out doesn't change
  If there is a node Y such that Y is not Z and is not in out and Y is conditionally dependent on Z given out, add Y to Out.
  
loop2
  if there is a Y in out such that Y is conditionally independent of Z given Out except Y
  remove Y from Out
  

## Hidden Markov models

```{r}



library(HMM)

states <- c("FD", "LD")
symbols <- c("1", "2", "3", "4", "5", "6")

start_probs <- rep(1/2, 2) 
trans_probs = matrix(c(.5, .5,
                        .5, .5
  
), nrow=length(states), ncol=length(states), byrow = TRUE)
colnames(trans_probs) = states
rownames(trans_probs) = states

TP = 1/6 
FP = 1/2
L = 1/10
emission_probs <- matrix(c(TP,TP,TP,TP,TP,TP,
                           L,L,L,L,L,FP
), nrow=length(states), ncol=length(symbols), byrow = TRUE)

colnames(emission_probs) = symbols
rownames(emission_probs) = states


# Initialize the Hidden Markov Model
hmm_model <- initHMM(
  States = states,           # vector of states
  Symbols = symbols,         # vector of observation symbols
  startProbs = start_probs,  # Initial state probabilities
  transProbs = trans_probs,  # Transition probabilities matrix
  emissionProbs = emission_probs  # Emission probabilities matrix
)

set.seed(12345)
sim = simHMM(hmm_model, length = 10000)

countLD = 0
countFD = 0
for (i in 1:10000) {
  if(sim$observation[i] == 6 && sim$states[i] == "LD"){
    countLD = countLD +1
  }else if(sim$observation[i] == 6 && sim$states[i] == "FD")
    countFD = countFD+1
}

countLD
countFD

```
Seems reasonable since loaded die get 6 more often

```{r}


library(HMM)

states <- c("FD1", "FD2", "FD3", "LD1", "LD2", "LD3")
symbols <- c("1", "2", "3", "4", "5", "6")

start_probs <- rep(1/6, 6) 
trans_probs = matrix(c(.5,.5,0,0,0,0,
                       0,.5,.5,0,0,0,
                       0,0,.5,.5,0,0,
                       0,0,0,.5,.5,0,
                       0,0,0,0,.5,.5,
                       .5,0,0,0,0,.5
), nrow=length(states), ncol=length(states), byrow = TRUE)
colnames(trans_probs) = states
rownames(trans_probs) = states

TP = 1/6 
FP = 1/2
LN = 1/10
emission_probs <- matrix(c(TP,TP,TP,TP,TP,TP,
                           TP,TP,TP,TP,TP,TP,
                           TP,TP,TP,TP,TP,TP,
                           LN,LN,LN,LN,LN,FP,
                           LN,LN,LN,LN,LN,FP,
                           LN,LN,LN,LN,LN,FP
), nrow=length(states), ncol=length(symbols), byrow = TRUE)

colnames(emission_probs) = symbols
rownames(emission_probs) = states


# Initialize the Hidden Markov Model
hmm_model <- initHMM(
  States = states,           # vector of states
  Symbols = symbols,         # vector of observation symbols
  startProbs = start_probs,  # Initial state probabilities
  transProbs = trans_probs,  # Transition probabilities matrix
  emissionProbs = emission_probs  # Emission probabilities matrix
)

set.seed(12345)
sim = simHMM(hmm_model, length = 10000)

countLD = 0
countFD = 0
for (i in 1:10000) {
  if(sim$observation[i] == 6 && (sim$states[i] == "LD1" || sim$states[i] == "LD2" || sim$states[i] == "LD3")){
    countLD = countLD +1
  }else if(sim$observation[i] == 6 && (sim$states[i] == "FD1" || sim$states[i] == "FD2" || sim$states[i] == "FD3"))
    countFD = countFD+1
}

countLD
countFD

```

